---
title: "PresentBayesian Regression for Spatial Rainfall Prediction and Analysis"
subtitle: "A Comparative Study with Linear Regression and Uncertainty Quantification"
author: "Mahmudul Islam Prakash (Advisor: Dr. Cohen)"
date: today
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction  {.smaller}

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.

-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

In kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]\*

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.



## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
